### 2017-12-20 chapters/paper4/main.tex
Fix labels and reference in the papers

\subsection{Mechanical modeling}
\label{sec:mechanical_modeling}
We represent the heart as a continuum body with a reference configuration 
$\Omega \subset \mathbb{R}^3$, and let $\Xvec$ denote the
corresponding reference coordinate. Further let $ \Omega_t
\subset \mathbb{R}^3$ be the deformed domain after a given time $t >
0$ and let $\xvec$ be the corresponding physical coordinate.  The
motion of a point $\Xvec \in  \Omega$ to the  point $\xvec \in
\Omega_t$ may be described by a deformation map  $\varphi :
\Omega  \rightarrow \Omega_t$.

@@

The myocardium is modeled as a hyperelastic material, and we adapt a
transversely isotropic law from \cite{holzapfel2009constitutive},
where the strain-energy function $\Psi$ is given by  
\begin{align}
  \label{eq:hoa}
  \Psi(\Cbar) = \frac{a}{2 b} \left( e^{ b (\bar{I}_1  - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (\bar{I}_{4\ef} - 1)_+^2} -1 \right). 
\end{align}
Here $\bar{I}_1 = \tr \Cbar$ and $\bar{I}_{4\ef}= \ef \cdot (\Cbar
\ef)$ with $\Cbar = J^{-2/3}\C$ being the isochoric part of the right
Cauchy-Green deformation tensor , $\ef$ is a
unit vector field in the direction of the myofibers, $(\cdotp)_{+} 
= \max\{\cdot, 0\}$, and $a, a_f, b, b_f$ are 
material stiffness parameters defining the elastic properties of the
myocardium.

@@

To model the active contraction of the ventricle we employ the active
strain formulation. In this formulation one assumes a multiplicative 
decomposition of the deformation gradient into an elastic
and an active component \cite{ambrosi2011electromechanical}:
\begin{equation}
 \F = \F_e \F_a.
\label{eq:active_strain}
\end{equation}
The active part, $\F_a$ represents the active shortening of the cells,
while the elastic part $\F_e$ ensures compatibility of the tissue.
Similar to \cite{balaban2017high} we choose the following form of the active
deformation gradient,
\begin{equation}
  \F_a = (1 - \gamma) \ef \otimes \ef  + \frac{1}{\sqrt{1 - \gamma}} (\I - \ef \otimes \ef).
 \label{eq:active_strain_Fa_gjerald}
\end{equation}
The active deformation gradient $\F_a$ does not contribute to any
stored energy within the material, so that the modified
strain-energy function $\tilde{\Psi}(\C_e) = \Psi(\F_a^{-T} \Cbar
\F_a^{-1})$, which is a pure function of the elastic deformations is used.

@@

To solve the problem we employ the Galerkin finite element method, which is based
on choosing proper finite dimensional subspaces $V_h \subset [H_0^1(\Omega)]^3$ and
$Q_h \subset L^2(\Omega)$ and solve the weak formulation of the problem. In
the Lagrangian form, the weak form reads: {\it find $(\uvec,p)
\in V_h\times Q_h$ such that for all variations $(\delta
\uvec, \delta p) \in V_h \times Q_h $ and $\left. \left( \uvec \cdot \N \right)
\right|_{\partial \Omega_{\mathrm{base}}} = 0$ }:
\begin{align}
\begin{split}
  F(\uvec, p;\delta \uvec, \delta p) =
  &\int_{\Omega} \FPiola :  \Grad \delta \uvec  \mathrm{d} V  \\
  &+\int_{\Omega} p J \F^{-T} : \Grad \ \delta \uvec   \mathrm{d} V  
  +\int_{\Omega} (J - 1) \delta p \mathrm{d} V \\
  &+\int_{\partial \Omega_{\text{base}}} k \uvec \cdot \N \mathrm{d}S
  - \int_{\partial \Omega_{\text{endo}}} \plv J \F^{-T} \N \cdot \delta \uvec \mathrm{d}S= 0.
  \label{eq:force_balance}
\end{split}
\end{align}
Here, the ventricular domain is anchored by constraining the basal
motion by a linear spring-term with stiffness $k$, and the
endocardial cavity pressure $\plv$ is enforced as a Neumann boundary
condition on the endocardium.

@@

\subsection{Calculation of work}
\label{sec:regional_work}

@@

\subsubsection{Quasi-static approximation}
The mechanical model described in Section
\ref{sec:mechanical_modeling} is quasi-static, meaning that the
inertial forces are assumed to be so small that the
equilibrium equation \eqref{eq:force_balance} holds at any discrete
time point. Therefore, strain rate is estimated using a
backward difference approximation \cite{wang2011myocardial}.
Suppose we have $N+1$ discrete time points $t_i, i = 0,1, \cdots N$, where $ t_0$
refers to the the reference configuration, and let $\SPiola(t, \Xvec)$
and $\E(t, \Xvec)$ be respectively the stress and strain tensors at
time $t$ and position $\Xvec$ in the reference configuration. In the
following, the full stress and strain tensors are used, but the methodology
also applies to any component of the stress and strain tensor. We define
\begin{align}
  \overline{\SPiola}(t_i,\Xvec) &= \frac{1}{2}\left( \mathbf{S}(t_i,\Xvec) + \mathbf{S}(t_{i-1},\Xvec) \right),\\
  d\E(t_i,\Xvec) &= \E(t_i,\Xvec) - \E(t_{i-1},\Xvec),
\end{align}
for $i = 1,\cdots, N$, and note that $ \mathbf{S}(t_0) = \mathbf{E}(t_0) = \mathbf{0}$.
The mechanical work in a region $\Omega_j$ from time $t_m$ to time $t_n$ can now be approximated
as 
\begin{align}
  \begin{split}
   \mathcal{W}(t_m, t_n)[\Omega_j]
  &= -\int_{t_m}^{t_n} \int_{\Omega_j}\SPiola(t, \Xvec) : \dot{\E}(t, \Xvec) \mathrm{d}\Xvec \mathrm{d}t \\
  &\approx - \sum_{i = m}^{n}\int_{\Omega_j}\overline{\SPiola}(t_i, \Xvec) : d\E(t_i, \Xvec) \mathrm{d}\Xvec
  \end{split}
\label{eq:total_work}
\end{align}
Similarly, the mechanical work computed from the pressure-volume loop area can be estimated as  
\begin{align}
  \begin{split}
  \mathcal{W}(t_m,t_n)[\Omega]
  &= - \int_{t_m}^{t_n} \frac{\mathrm{d}V}{\mathrm{d}t} P(t) \mathrm{d}t \\
  &\approx -\sum_{i= m}^{n}  (V(t_i) - V(t_{i-1})) \cdot \frac{P(t_i) + P(t_{i-1})}{2}.
  \end{split}
\end{align}

\subsubsection{Efficiency and wasted work ratio (WWR)}
\label{sec:efficiency_wwr}

During systole, most segments shorten and perform positive work
($\mathcal{W}_{\mathrm{pos}}$). However some segments might elongate,
resulting in energy loss, and we say that these segments do negative work
($\mathcal{W}_{\mathrm{neg}}$). Positive work is identified as the points with
$d\E_{\ef}(t) < 0$ (shortening), while negative work is identified as points with
$d\E_{\ef}(t) > 0$ (lengthening).
A local measure of the energetic efficacy of each segment is the ratio between the
negative and positive work, an index known as the segmental wasted
work ratio (WWR) \cite{russell2013assessment},
\begin{align}
  \mathrm{WWR} = \frac{|\mathcal{W}_{\mathrm{neg}}|}{\mathcal{W}_{\mathrm{pos}}}
\end{align}
Another measure of efficiency, formulated by Niederer et al
\cite{niederer2009role} is the ratio of positive work and the total
work
\begin{align}
  \eta = \frac{\mathcal{W}_{\mathrm{pos}}}{\mathcal{W}_{\mathrm{pos}} + |\mathcal{W}_{\mathrm{neg}}|}.
  \label{eq:efficiency}
\end{align}

The two indices are related by $\eta = \left( 1 +
  \mathrm{WWR} \right)^{-1}$, and we will therefore only report values
of the efficiency \eqref{eq:efficiency}, for which a value close to 1.0 indicate that most
of the performed work is positive. Since we expect work during systole
to be the main contribution to the overall ejection of blood, the
index is computed by only including work done from mitral valve closing to
mitral valve opening \cite{russell2013assessment}.

@@

In order to constrain the mechanical model to clinical data, we
applied a variational data assimilation technique. Let $\mathbf{w} =
(\uvec,p)$ denote a joint state variable consisting of the
displacement field $\uvec$ and the hydrostatic pressure $p$. The LV
cavity volume can be approximated as  
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\mathbf{w})= -\frac{1}{3}\int_{\lvendo}
  \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S, 
  \label{eq:volume_operator}
\end{align}
where $\N$ denotes the unit facet normal. Furthermore, the average regional
longitudinal strain can be approximated in region number $j$ as
\begin{align}
  \mathcal{H}_{\mathrm{strain}, j}(\mathbf{w}) = \frac{1}{|\Omega_j|}\int_{\Omega_j}
  \mathbf{e}_l^T \left( \mathbf{\tilde{\F} - \I} \right)\mathbf{e}_l  \mathrm{d}V.
\end{align}
Here $\mathbf{e}_l$ denotes a unit vector field in
the longitudinal direction, and $\tilde{\F}$ refers to the modified deformation
gradient relative to the reference geometry in the measurements.

For a given measurement $y_k$, we want to find a state variable
$\mathbf{w}$ that minimizes the mismatch between $y_k$ and
$\mathcal{H}_k(\mathbf{w})$. The state variable is determined by
tuning model-based parameters, also
referred to as control parameters, which we generically denote
by $\mu$. This can be formulated as a PDE-constrained optimization
problem of the form
\begin{equation}
  \begin{aligned}
    \label{eq:pde_opt}
    & \underset{\mu}{\text{minimize}}
    & &  \mathcal{J}(\mathbf{w}(\mu), \mu) \\
    & \text{subject to}
    & & F(\mathbf{w}, \mu)  = 0,
  \end{aligned}
\end{equation}
where $F=0$ represents the PDE-constraint given by
\eqref{eq:force_balance}, and $\mathcal{J}$ is a possibly
multiobjective function  
\begin{align}
  \mathcal{J}(\mathbf{w}(\mu), \mu) = \sum_{k = 1}^{N} w_k j_k (\mathbf{w}(\mu), \mu),
  \label{eq:cost_functional}
\end{align}
with $w_k$ being scalar weights, and $j_k = d( y_k,
  \mathcal{H}_k(\mathbf{w}) ) $ for some suitable metric $d$.

\subsubsection{Joint estimation of unloaded configuration and passive
  material parameters}
\label{sec:passive_param_estim}

One challenge when using image-based geometries is that the geometry
observed in a medical image is subjected to loads, e.g blood
pressure, while our finite element formulation assumes a stress-free
reference configuration. Therefore we need to find a geometry, so
that when the measured pressure in the image is applied, we recover
the image-based geometry. Several methods for estimating the unloaded,
zero-pressure geometry exist, including the inverse design analysis
(ID) \cite{govindjee1996computational}, the modified updated
Lagrangian formulation (MULF) \cite{gee2010computational} and
iterative methods \cite{bols2013computational,Rausch2017JB}.
Here we apply a simple fixed-point iteration scheme known as the backward
displacement method \cite{bols2013computational}.
The estimated unloaded configuration will depend upon the choice of material
parameters, which is not know  \emph{a priori}. In Appendix
\ref{sec:unloading} we explain the main steps in the backward
displacement method and a proposed method for joint estimation of material
parameters unloaded configuration. Additionally, an analysis of
the convergence properties of this method is also provided.

\subsubsection{Active parameter estimation}
\label{sec:active_param_estim}
From the first measurement point after end-diastole and throughout the
remaining cardiac cycle, that parameter $\gamma$ in
\eqref{eq:active_strain_Fa_gjerald} is used as control in
\eqref{eq:pde_opt}, and the cost functional \eqref{eq:cost_functional}
at time point $i$ is given by 
\begin{align}
  \mathcal{J} =
   \left(\frac{\mathcal{H}_{\mathrm{volume}}(\uvec^i)
  - V^i}{V^i} \right)^2 +
  \sum_{j=1}^{12} \delta_j \left( \mathcal{H}_{\mathrm{strain}, j}(\uvec^i)
  - \varepsilon_{j}^i \right)^2  + \lambda \| \nabla \gamma \|^2.
\end{align}
Here $V^i$ is the measured cavity volume at time point $i$, 
$\varepsilon_{j}^i$ is the measured longitudinal strain in segment $j$
at time point $i$, and $\delta_j$ is 0 if the measurement in segment
$j$ is missing and 1 otherwise. Note that even though the  
AHA-representation \cite{cerqueira2002standardized} contains 17
segments, we only include the mid and basal segments in the
optimization because of the relatively poor quality of the apical
segments, leaving us with a total of 12 segments.

@@

All simulations were carried out using an in-house software
package\footnote{Source code is publicly available at
  \url{https://bitbucket.org/finsberg/pulse_adjoint}} 
based on FEniCS \cite{logg2012automated} and Dolfin-Adjoint \cite{farrell2013automated}.
In these simulations the value of the linear spring $k$ in
\eqref{eq:force_balance} was determined by testing a range of values,
and choosing the softest spring that allowed for a convergent
estimation of the unloaded configuration. Based on this experiment we
choose $k=0.01$ kPa/cm$^2$.

During the estimation of the active and passive parameters outlined in Section
\ref{sec:active_param_estim} and \ref{sec:passive_param_estim}
respectively, the optimization was terminated if the difference between two
successive iterations was less then $10^{-6}$ or the number of
iterations exceeded 100. For the material parameter, an upper and lower
bound was set to $50$ and $0.05$ kPa respectively, while for the active
parameter these bounds were set to $0.4$ and $0.0$, allowing for up to
40 $\%$ active shortening of the myofibers.

@@

% The linear isotropic material parameter $a$ in \eqref{eq:hoa}, along
% with an unloaded configuration was estimated by means of Algorithm
% \ref{alg:unload_mat_bdm} for both the normal and the LBBB case.
% Figure \ref{fig:unloaded_simulation} shows the value of the material
% parameters, and the residual representing the relative difference
% between the unloaded cavity volumes of two successive iterations.
% The LBBB and normal case used respectively
% $7$ and $17$ iterations to get the residual below 0.01, resulting in a
% material parameters of $a= 32.1$ kPa for the LBBB case and $a= 11.96$ kPa for the
% normal case.
% \begin{figure}[htbp]
%   \centering
%   \includegraphics{figures/unloaded_simulation}
%   \caption{\label{fig:unloaded_simulation}Estimation of material
%     parameters for normal (blue) and LBBB (red) case. Solid lines
%     show the value of the material parameter at each iteration with
%   initial guess in both cases being $a=1.291$ kPa (iteration -1). Then
% the offset of the pressure in the image-based geometry was subtracted
% and new material parameters was estimated (iteration 0), and Algorithm
% \ref{alg:unload_mat_bdm} was applied using this estimate as initial
% guess. Dashed lines are the residuals, representing the relative
% difference in unloaded cavity volume. The algorithm terminated when
% the residual was below 1 \%. }
% \end{figure}
% Figure \ref{fig:unloaded_geo} shows the resulting
% unloaded configurations on top of the original geometries. 
% \begin{figure}[htbp]
%   \centering
%   \includegraphics{figures/unloading}
%   \caption{\label{fig:unloaded_geo}Estimated unloaded
%     configuration for normal (right) and LBBB (left) case in red, and
%     the original image-based geometry in transparent. }
% \end{figure}

@@

The simulated and measured PV loops and the regional
pressure-strain loops are shown in Figure \ref{fig:pv_loops} and
\ref{fig:regional_strain_pressure_loops} respectively. 
For the normal case, the longitudinal strain curves in the basal
inferior and posterior segments were missing, and so only the
simulated results are shown in these segments.

\begin{figure}[htbp]
  \centering
  \includegraphics{figures/pv_loop}
  \caption{\label{fig:pv_loops}Simulated and measured PV loops
    for the LBBB (left) and normal (right) case. }
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics{figures/pressure_strain_loop}
\caption{\label{fig:regional_strain_pressure_loops}Measured and
simulated pressure-strain loops for LBBB (top) and normal (bottom)
case. Strain measurement in basal inferior and posterior segments for
the normal subject were discarded in the optimization due to poor
quality, and therefore only simulated results are shown in these
segments.} 
\end{figure}

@@

Analogous to the regional pressure-strain loops in Figure
\ref{fig:regional_strain_pressure_loops}, the resulting regional fiber
stress-strain loops are depicted in Figure
\ref{fig:regional_fiber_stress_strain_loops}. Here we show Cauchy
fiber stress on the $y-$axis and for the sake of correct illustration
of regional work in terms of loop area, we show the Almansi fiber
strain on the $x-$axis.

\begin{figure}[htbp]
  \centering
  \includegraphics{figures/fiber_stress_strain_loop}
  \caption{\label{fig:regional_fiber_stress_strain_loops}Simulated
    regional fiber stress-strain loops in the mid and basal segments
    for normal (blue) and LBBB (red). } 
\end{figure}

The Cauchy fiber stress visualized at the deformed configurations at
the different valvular events for the normal and LBBB case are shown
in Figure \ref{fig:snap_shots}. The LBBB case shows clear signs of
dyssynchrony and more uneven distribution of stress compared to the
normal case.

>>>>

\subsection{Mechanical modeling}
\label{paper4:sec:mechanical_modeling}
We represent the heart as a continuum body with a reference configuration 
$\Omega \subset \mathbb{R}^3$, and let $\Xvec$ denote the
corresponding reference coordinate. Further let $ \Omega_t
\subset \mathbb{R}^3$ be the deformed domain after a given time $t >
0$ and let $\xvec$ be the corresponding physical coordinate.  The
motion of a point $\Xvec \in  \Omega$ to the  point $\xvec \in
\Omega_t$ may be described by a deformation map  $\varphi :
\Omega  \rightarrow \Omega_t$.

@@

The myocardium is modeled as a hyperelastic material, and we adapt a
transversely isotropic law from \cite{holzapfel2009constitutive},
where the strain-energy function $\Psi$ is given by  
\begin{align}
  \label{paper4:eq:hoa}
  \Psi(\Cbar) = \frac{a}{2 b} \left( e^{ b (\bar{I}_1  - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (\bar{I}_{4\ef} - 1)_+^2} -1 \right). 
\end{align}
Here $\bar{I}_1 = \tr \Cbar$ and $\bar{I}_{4\ef}= \ef \cdot (\Cbar
\ef)$ with $\Cbar = J^{-2/3}\C$ being the isochoric part of the right
Cauchy-Green deformation tensor , $\ef$ is a
unit vector field in the direction of the myofibers, $(\cdotp)_{+} 
= \max\{\cdot, 0\}$, and $a, a_f, b, b_f$ are 
material stiffness parameters defining the elastic properties of the
myocardium.

@@

To model the active contraction of the ventricle we employ the active
strain formulation. In this formulation one assumes a multiplicative 
decomposition of the deformation gradient into an elastic
and an active component \cite{ambrosi2011electromechanical}:
\begin{equation}
 \F = \F_e \F_a.
\label{paper4:eq:active_strain}
\end{equation}
The active part, $\F_a$ represents the active shortening of the cells,
while the elastic part $\F_e$ ensures compatibility of the tissue.
Similar to \cite{balaban2017high} we choose the following form of the active
deformation gradient,
\begin{equation}
  \F_a = (1 - \gamma) \ef \otimes \ef  + \frac{1}{\sqrt{1 - \gamma}} (\I - \ef \otimes \ef).
 \label{paper4:eq:active_strain_Fa_gjerald}
\end{equation}
The active deformation gradient $\F_a$ does not contribute to any
stored energy within the material, so that the modified
strain-energy function $\tilde{\Psi}(\C_e) = \Psi(\F_a^{-T} \Cbar
\F_a^{-1})$, which is a pure function of the elastic deformations is used.

@@

To solve the problem we employ the Galerkin finite element method, which is based
on choosing proper finite dimensional subspaces $V_h \subset [H_0^1(\Omega)]^3$ and
$Q_h \subset L^2(\Omega)$ and solve the weak formulation of the problem. In
the Lagrangian form, the weak form reads: {\it find $(\uvec,p)
\in V_h\times Q_h$ such that for all variations $(\delta
\uvec, \delta p) \in V_h \times Q_h $ and $\left. \left( \uvec \cdot \N \right)
\right|_{\partial \Omega_{\mathrm{base}}} = 0$ }:
\begin{align}
\begin{split}
  F(\uvec, p;\delta \uvec, \delta p) =
  &\int_{\Omega} \FPiola :  \Grad \delta \uvec  \mathrm{d} V  \\
  &+\int_{\Omega} p J \F^{-T} : \Grad \ \delta \uvec   \mathrm{d} V  
  +\int_{\Omega} (J - 1) \delta p \mathrm{d} V \\
  &+\int_{\partial \Omega_{\text{base}}} k \uvec \cdot \N \mathrm{d}S
  - \int_{\partial \Omega_{\text{endo}}} \plv J \F^{-T} \N \cdot \delta \uvec \mathrm{d}S= 0.
  \label{paper4:eq:force_balance}
\end{split}
\end{align}
Here, the ventricular domain is anchored by constraining the basal
motion by a linear spring-term with stiffness $k$, and the
endocardial cavity pressure $\plv$ is enforced as a Neumann boundary
condition on the endocardium.

@@

\subsection{Calculation of work}
\label{paper4:sec:regional_work}

@@

\subsubsection{Quasi-static approximation}
The mechanical model described in Section
\ref{paper4:sec:mechanical_modeling} is quasi-static, meaning that the
inertial forces are assumed to be so small that the
equilibrium equation \eqref{paper4:eq:force_balance} holds at any discrete
time point. Therefore, strain rate is estimated using a
backward difference approximation \cite{wang2011myocardial}.
Suppose we have $N+1$ discrete time points $t_i, i = 0,1, \cdots N$, where $ t_0$
refers to the the reference configuration, and let $\SPiola(t, \Xvec)$
and $\E(t, \Xvec)$ be respectively the stress and strain tensors at
time $t$ and position $\Xvec$ in the reference configuration. In the
following, the full stress and strain tensors are used, but the methodology
also applies to any component of the stress and strain tensor. We define
\begin{align}
  \overline{\SPiola}(t_i,\Xvec) &= \frac{1}{2}\left( \mathbf{S}(t_i,\Xvec) + \mathbf{S}(t_{i-1},\Xvec) \right),\\
  d\E(t_i,\Xvec) &= \E(t_i,\Xvec) - \E(t_{i-1},\Xvec),
\end{align}
for $i = 1,\cdots, N$, and note that $ \mathbf{S}(t_0) = \mathbf{E}(t_0) = \mathbf{0}$.
The mechanical work in a region $\Omega_j$ from time $t_m$ to time $t_n$ can now be approximated
as 
\begin{align}
  \begin{split}
   \mathcal{W}(t_m, t_n)[\Omega_j]
  &= -\int_{t_m}^{t_n} \int_{\Omega_j}\SPiola(t, \Xvec) : \dot{\E}(t, \Xvec) \mathrm{d}\Xvec \mathrm{d}t \\
  &\approx - \sum_{i = m}^{n}\int_{\Omega_j}\overline{\SPiola}(t_i, \Xvec) : d\E(t_i, \Xvec) \mathrm{d}\Xvec
  \end{split}
\label{paper4:eq:total_work}
\end{align}
Similarly, the mechanical work computed from the pressure-volume loop area can be estimated as  
\begin{align}
  \begin{split}
  \mathcal{W}(t_m,t_n)[\Omega]
  &= - \int_{t_m}^{t_n} \frac{\mathrm{d}V}{\mathrm{d}t} P(t) \mathrm{d}t \\
  &\approx -\sum_{i= m}^{n}  (V(t_i) - V(t_{i-1})) \cdot \frac{P(t_i) + P(t_{i-1})}{2}.
  \end{split}
\end{align}

\subsubsection{Efficiency and wasted work ratio (WWR)}
\label{paper4:sec:efficiency_wwr}

During systole, most segments shorten and perform positive work
($\mathcal{W}_{\mathrm{pos}}$). However some segments might elongate,
resulting in energy loss, and we say that these segments do negative work
($\mathcal{W}_{\mathrm{neg}}$). Positive work is identified as the points with
$d\E_{\ef}(t) < 0$ (shortening), while negative work is identified as points with
$d\E_{\ef}(t) > 0$ (lengthening).
A local measure of the energetic efficacy of each segment is the ratio between the
negative and positive work, an index known as the segmental wasted
work ratio (WWR) \cite{russell2013assessment},
\begin{align}
  \mathrm{WWR} = \frac{|\mathcal{W}_{\mathrm{neg}}|}{\mathcal{W}_{\mathrm{pos}}}
\end{align}
Another measure of efficiency, formulated by Niederer et al
\cite{niederer2009role} is the ratio of positive work and the total
work
\begin{align}
  \eta = \frac{\mathcal{W}_{\mathrm{pos}}}{\mathcal{W}_{\mathrm{pos}} + |\mathcal{W}_{\mathrm{neg}}|}.
  \label{paper4:eq:efficiency}
\end{align}

The two indices are related by $\eta = \left( 1 +
  \mathrm{WWR} \right)^{-1}$, and we will therefore only report values
of the efficiency \eqref{paper4:eq:efficiency}, for which a value close to 1.0 indicate that most
of the performed work is positive. Since we expect work during systole
to be the main contribution to the overall ejection of blood, the
index is computed by only including work done from mitral valve closing to
mitral valve opening \cite{russell2013assessment}.

@@

In order to constrain the mechanical model to clinical data, we
applied a variational data assimilation technique. Let $\mathbf{w} =
(\uvec,p)$ denote a joint state variable consisting of the
displacement field $\uvec$ and the hydrostatic pressure $p$. The LV
cavity volume can be approximated as  
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\mathbf{w})= -\frac{1}{3}\int_{\lvendo}
  \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S, 
  \label{paper4:eq:volume_operator}
\end{align}
where $\N$ denotes the unit facet normal. Furthermore, the average regional
longitudinal strain can be approximated in region number $j$ as
\begin{align}
  \mathcal{H}_{\mathrm{strain}, j}(\mathbf{w}) = \frac{1}{|\Omega_j|}\int_{\Omega_j}
  \mathbf{e}_l^T \left( \mathbf{\tilde{\F} - \I} \right)\mathbf{e}_l  \mathrm{d}V.
\end{align}
Here $\mathbf{e}_l$ denotes a unit vector field in
the longitudinal direction, and $\tilde{\F}$ refers to the modified deformation
gradient relative to the reference geometry in the measurements.

For a given measurement $y_k$, we want to find a state variable
$\mathbf{w}$ that minimizes the mismatch between $y_k$ and
$\mathcal{H}_k(\mathbf{w})$. The state variable is determined by
tuning model-based parameters, also
referred to as control parameters, which we generically denote
by $\mu$. This can be formulated as a PDE-constrained optimization
problem of the form
\begin{equation}
  \begin{aligned}
    \label{paper4:eq:pde_opt}
    & \underset{\mu}{\text{minimize}}
    & &  \mathcal{J}(\mathbf{w}(\mu), \mu) \\
    & \text{subject to}
    & & F(\mathbf{w}, \mu)  = 0,
  \end{aligned}
\end{equation}
where $F=0$ represents the PDE-constraint given by
\eqref{paper4:eq:force_balance}, and $\mathcal{J}$ is a possibly
multiobjective function  
\begin{align}
  \mathcal{J}(\mathbf{w}(\mu), \mu) = \sum_{k = 1}^{N} w_k j_k (\mathbf{w}(\mu), \mu),
  \label{paper4:eq:cost_functional}
\end{align}
with $w_k$ being scalar weights, and $j_k = d( y_k,
  \mathcal{H}_k(\mathbf{w}) ) $ for some suitable metric $d$.

\subsubsection{Joint estimation of unloaded configuration and passive
  material parameters}
\label{paper4:sec:passive_param_estim}

One challenge when using image-based geometries is that the geometry
observed in a medical image is subjected to loads, e.g blood
pressure, while our finite element formulation assumes a stress-free
reference configuration. Therefore we need to find a geometry, so
that when the measured pressure in the image is applied, we recover
the image-based geometry. Several methods for estimating the unloaded,
zero-pressure geometry exist, including the inverse design analysis
(ID) \cite{govindjee1996computational}, the modified updated
Lagrangian formulation (MULF) \cite{gee2010computational} and
iterative methods \cite{bols2013computational,Rausch2017JB}.
Here we apply a simple fixed-point iteration scheme known as the backward
displacement method \cite{bols2013computational}.
The estimated unloaded configuration will depend upon the choice of material
parameters, which is not know  \emph{a priori}. In Appendix
\ref{paper4:sec:unloading} we explain the main steps in the backward
displacement method and a proposed method for joint estimation of material
parameters unloaded configuration. Additionally, an analysis of
the convergence properties of this method is also provided.

\subsubsection{Active parameter estimation}
\label{paper4:sec:active_param_estim}
From the first measurement point after end-diastole and throughout the
remaining cardiac cycle, that parameter $\gamma$ in
\eqref{paper4:eq:active_strain_Fa_gjerald} is used as control in
\eqref{paper4:eq:pde_opt}, and the cost functional \eqref{paper4:eq:cost_functional}
at time point $i$ is given by 
\begin{align}
  \mathcal{J} =
   \left(\frac{\mathcal{H}_{\mathrm{volume}}(\uvec^i)
  - V^i}{V^i} \right)^2 +
  \sum_{j=1}^{12} \delta_j \left( \mathcal{H}_{\mathrm{strain}, j}(\uvec^i)
  - \varepsilon_{j}^i \right)^2  + \lambda \| \nabla \gamma \|^2.
\end{align}
Here $V^i$ is the measured cavity volume at time point $i$, 
$\varepsilon_{j}^i$ is the measured longitudinal strain in segment $j$
at time point $i$, and $\delta_j$ is 0 if the measurement in segment
$j$ is missing and 1 otherwise. Note that even though the  
AHA-representation \cite{cerqueira2002standardized} contains 17
segments, we only include the mid and basal segments in the
optimization because of the relatively poor quality of the apical
segments, leaving us with a total of 12 segments.

@@

All simulations were carried out using an in-house software
package\footnote{Source code is publicly available at
  \url{https://bitbucket.org/finsberg/pulse_adjoint}} 
based on FEniCS \cite{logg2012automated} and Dolfin-Adjoint \cite{farrell2013automated}.
In these simulations the value of the linear spring $k$ in
\eqref{paper4:eq:force_balance} was determined by testing a range of values,
and choosing the softest spring that allowed for a convergent
estimation of the unloaded configuration. Based on this experiment we
choose $k=0.01$ kPa/cm$^2$.

During the estimation of the active and passive parameters outlined in Section
\ref{paper4:sec:active_param_estim} and \ref{paper4:sec:passive_param_estim}
respectively, the optimization was terminated if the difference between two
successive iterations was less then $10^{-6}$ or the number of
iterations exceeded 100. For the material parameter, an upper and lower
bound was set to $50$ and $0.05$ kPa respectively, while for the active
parameter these bounds were set to $0.4$ and $0.0$, allowing for up to
40 $\%$ active shortening of the myofibers.

@@

% The linear isotropic material parameter $a$ in \eqref{paper4:eq:hoa}, along
% with an unloaded configuration was estimated by means of Algorithm
% \ref{paper4:alg:unload_mat_bdm} for both the normal and the LBBB case.
% Figure \ref{paper4:fig:unloaded_simulation} shows the value of the material
% parameters, and the residual representing the relative difference
% between the unloaded cavity volumes of two successive iterations.
% The LBBB and normal case used respectively
% $7$ and $17$ iterations to get the residual below 0.01, resulting in a
% material parameters of $a= 32.1$ kPa for the LBBB case and $a= 11.96$ kPa for the
% normal case.
% \begin{figure}[htbp]
%   \centering
%   \includegraphics{figures/unloaded_simulation}
%   \caption{\label{paper4:fig:unloaded_simulation}Estimation of material
%     parameters for normal (blue) and LBBB (red) case. Solid lines
%     show the value of the material parameter at each iteration with
%   initial guess in both cases being $a=1.291$ kPa (iteration -1). Then
% the offset of the pressure in the image-based geometry was subtracted
% and new material parameters was estimated (iteration 0), and Algorithm
% \ref{paper4:alg:unload_mat_bdm} was applied using this estimate as initial
% guess. Dashed lines are the residuals, representing the relative
% difference in unloaded cavity volume. The algorithm terminated when
% the residual was below 1 \%. }
% \end{figure}
% Figure \ref{paper4:fig:unloaded_geo} shows the resulting
% unloaded configurations on top of the original geometries. 
% \begin{figure}[htbp]
%   \centering
%   \includegraphics{figures/unloading}
%   \caption{\label{paper4:fig:unloaded_geo}Estimated unloaded
%     configuration for normal (right) and LBBB (left) case in red, and
%     the original image-based geometry in transparent. }
% \end{figure}

@@

The simulated and measured PV loops and the regional
pressure-strain loops are shown in Figure \ref{paper4:fig:pv_loops} and
\ref{paper4:fig:regional_strain_pressure_loops} respectively. 
For the normal case, the longitudinal strain curves in the basal
inferior and posterior segments were missing, and so only the
simulated results are shown in these segments.

\begin{figure}[htbp]
  \centering
  \includegraphics{figures/pv_loop}
  \caption{\label{paper4:fig:pv_loops}Simulated and measured PV loops
    for the LBBB (left) and normal (right) case. }
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics{figures/pressure_strain_loop}
\caption{\label{paper4:fig:regional_strain_pressure_loops}Measured and
simulated pressure-strain loops for LBBB (top) and normal (bottom)
case. Strain measurement in basal inferior and posterior segments for
the normal subject were discarded in the optimization due to poor
quality, and therefore only simulated results are shown in these
segments.} 
\end{figure}

@@

Analogous to the regional pressure-strain loops in Figure
\ref{paper4:fig:regional_strain_pressure_loops}, the resulting regional fiber
stress-strain loops are depicted in Figure
\ref{paper4:fig:regional_fiber_stress_strain_loops}. Here we show Cauchy
fiber stress on the $y-$axis and for the sake of correct illustration
of regional work in terms of loop area, we show the Almansi fiber
strain on the $x-$axis.

\begin{figure}[htbp]
  \centering
  \includegraphics{figures/fiber_stress_strain_loop}
  \caption{\label{paper4:fig:regional_fiber_stress_strain_loops}Simulated
    regional fiber stress-strain loops in the mid and basal segments
    for normal (blue) and LBBB (red). } 
\end{figure}

The Cauchy fiber stress visualized at the deformed configurations at
the different valvular events for the normal and LBBB case are shown
in Figure \ref{paper4:fig:snap_shots}. The LBBB case shows clear signs of
dyssynchrony and more uneven distribution of stress compared to the
normal case.

\begin{figure}[htbp]
  \centering
  \includegraphics{figures/snap_shots}
  \caption{\label{paper4:fig:snap_shots}Cauchy fiber stress for the normal and LBBB case at
    different valvular events (MVC: mitral valve closing, AVO: aortic
    valve opening, AVC: aortic valve closing, MVO: mitral valve
    opening), visualized on the deformed configurations. For reference
    we show the 17 AHA-zone delineation of the left ventricle in the
    two cases in the bottom panel. From left to right we have the
    lateral, anterior, anteroseptal and septal segments. }
\end{figure}
